### 2020-02-04 UbiCompSubmission/bp_discussion.tex
addressing rebuttal bullets

Currently, the cost of biometric sensors and necessary infrastructure,
such as automated light and sound systems for adjusting the
environment, makes our approach most appropriate for high-value
workspaces, such as control rooms, command centers, or dispatch
offices. However, as standard office settings become more
personalizable (e.g., via adjustable desks, lighting, and sound
showers) and sensor costs decrease, our approach could be applied to
any office environment, and thus could impact a large percentage of
modern workers. As in modern cars, temperature and lighting could be
regulated on a per-person basis, which would allow the environment to
react to the person's current state and to maximize each person's
preferences and productivity (e.g., preferences of men and women in
temperature~\cite{Karjalainen07}). Further research would be needed to
identify how to balance needs across a group of office workers and how
to handle conflicting levels between different group members.

@@

\subsection{The Effect of Tai Chi on Stress}
As described in our study procedure, we offered and asked participants to attend a one-hour Tai Chi class per week for the last four weeks of the study. We recorded each participant's attendance, including which day they attended the offered classes. The two primary reasons for this intervention was to keep participants motivated to continue the self-reports over the long study period, and to offer a technique that might help reduce stress.

>>>>

Currently, the cost of biometric sensors and necessary infrastructure,
such as automated light and sound systems for adjusting the
environment, makes our approach most appropriate for high-value
workspaces, such as control rooms, command centers, or dispatch
offices. However, as standard office settings become more
personalizable (e.g., via adjustable desks, lighting, and sound
showers) and sensor costs decrease, our approach could be applied to
any office environment, and thus could impact a large percentage of
modern workers. As in modern cars, temperature and lighting could be
regulated on a per-person basis, which would allow the environment to
react to the person's current state and to maximize each person's
preferences and productivity (e.g., preferences of men and women in
temperature~\cite{Karjalainen07}). We believe that our results present a good step to more in situ usage of biometric sensing. Future studies can build on the evidence, and for example, reduce the effort required for the experience sampling from continuously rating stress levels to using biometrics as a ground truth with occasional validation of the predicted stress levels, or
identify how to balance needs across a group of office workers and how
to handle conflicting levels between different group members.

@@

\subsection{The Effect of Tai Chi on Stress}
As described in our study procedure,
due to our focus on stress, we offered participants an opportunity to be part of Thai Chi classes.
The two primary reasons for this intervention was to keep participants motivated to continue the self-reports over the long study period, and to offer a technique that might help reduce stress.
After consulting with other researchers, we decided to offer Tai Chi in the last month of the study, leaving the first month of the study unaltered. While participants were not required to take a Tai Chi class, they all took one per week except for two participants that opted out of the last two weeks of classes. Four participants explicitly stated that they liked the Tai Chi sessions or that they were “excellent”.
These classes were performed once per week with a duration of one hour.

### 2020-02-04 UbiCompSubmission/bp_intro_v4.tex
addressing rebuttal bullets

%!TEX root = bioPrediction_main.tex
\section{Introduction}
\textit{`The most valuable asset of a 21st-century institution (whether 
business or non-business) will be its knowledge workers and their 
productivity'}~\cite{drucker1999knowledge}. Knowledge workers constantly face 
challenges, such as a high work fragmentation, continuous disruptions and 
distractions, highly complex and demanding tasks, and long working 
hours~\cite{gonzalez2004constant,mark2008cost,czerwinski04diary}. 
Consequently, stress in the workplace is an ever-growing concern. It leads to 
fatigue, burnout and various other illnesses, ultimately resulting in work 
absences and marked productivity 
losses~\cite{hockey1997stress,setz2010stress,wrs2010}.
But while the effects of stress are obvious the causes can be more subtle. Stress can be caused by
a multitude of factors within the office including but not limited to workload, management issues, workplace harassment and bullying,
constant interruptions, and noisy environments. It could also be the result of external stressors seeping into the workplace.
Regardless of the cause, garnering an understanding of what stress looks like in the workplace could
prove invaluable to helping us develop effective interventions.
Likewise, being able to continuously measure stress in the workplace bears huge potential for better supporting 
knowledge workers in combatting workplace challenges, and improving productivity 
and well-being.

At the same time, these 
challenges that contribute to stress levels can make it hard for knowledge workers to stay focused, impact interaction with their digital environment, degrade the quality of work performed and hinder productivity.
For instance, the continuous disruptions that knowledge workers face in office work environments can lead to a higher error rate and lower 
performance~\cite{bailey2001effects,mark2008cost}.

Fortunately, with recent advances in sensing technologies there is increasing 
hope that we can accurately and automatically measure stress and it's related aspects. A 
number of HCI studies have discussed and investigated the use of biometrics to 
measure aspects such as a person's cognitive state 
%This change is regarding Reviewer 2's concern: what are the "aspects"?
(e.g., processing load)~\cite{beatty82,Kramer90,richter1998psychophysiological,Rowe98,Wilson2002}, 
and more specifically to measure stress~\cite{dishman2000stress,Setz2010}, 
or awakeness / (energetic) arousal~\cite{exler2016tired,mcduff2012affectaura,haag2004emotion}. Similarly,
studies have examined the occurrence, cause and effect of similar 
aspects in various contexts~\cite{mark2014bored,mark2008cost}, how these can be measured~\cite{Hernandez14}, and 
 how they might be modulated/regulated~\cite{Snyder15}. 
While previous work mostly focused on smaller or controlled lab studies, our study extends this work by performing a field study to examine whether and how these three variables ---stress, focus and awakeness--- can be measured over an extended period of time in a knowledge worker's context, 
%Yet, very few 
%studies measure productivity-related factors in the workplace over an 
%extended period of time
and little is known about measuring the more 
abstract concept of focus (i.e., directed attention). 
% Added regarding concern of Reviewer 2: Define what is meant by focus when it's first introduced. 
Given the importance of focus, described as a 
combination of engagement and challenge in a work task by Mark et al.~\cite{mark2014bored}, to an office workers' productivity, this factor also 
deserves further study.

In this research we examine what office stress actually looks like over long periods of time, and develop continuous (and automatic) 
measures of stress, focus, and awakeness in the workplace. These measures are targeted towards improving knowledge workers' productivity and 
well-being in the future by customizing interaction with their environment; e.g., by continuously monitoring stress levels and recommending stress-reducing interventions during 
extended periods of high stress, automatically adjusting lighting and temperature to heighten awakeness, or improving management 
of disruptions~\cite{iqbal2005effectiveness,zuger2017reducing} to aid with focus and productivity.

Our work builds upon previous work and extends it by performing an eight-week 
study in the workplace with 14 knowledge workers, collecting biometric and 
computer interaction data as well as experience samples on focus, awakeness, 
and stress. Our participants, who perform various job functions for a research 
and development group within a single large corporation, wore a single 
biometric armband sensor\footnote{Biovotion Everion sensor~\cite{everion}} 
with low invasiveness that captured heart-, respiration- and skin-related measures. This 
modality was chosen to ease longitudinal deployment in the field. Using 
machine learning, we created classifiers and analyzed their ability to predict stress, focus, and awakeness levels.
%Similarly, we captured computer 
%interaction features from participants' everyday computer usage and
%analyzed how a model trained using the combination
%of these two training sets compares to the performance of models using each
%set individually.

In our analysis we identify several trends in day-to-day stress levels that emerged over the course of our eight-week study.
The results of our analysis also show that biometric signals collected from a single minimally invasive sensor can be used to predict stress and its related aspects accurately, with the abstract concept of focus (predictably) being the hardest to detect. Our results further show that knowledge workers' self-reported levels of stress, focus and awakeness and their physiological manifestation and prediction can vary a lot between individuals. 
%Our results further show that overall combining biometric features with computer interaction information 
%results in an improvement
%over each of the training sets individually.
The main contributions of our work are:
\begin{itemize}
	\item An qualitative examination of how knowledge worker stress presents and fluctuates in the wild.
	\item The creation and analysis of measures for the automatic monitoring of knowledge workers' awakeness, stress, and focus in the workplace based on an eight-week field study with 14 office workers.
	%\item A thorough analysis of ways to increase performance, including the training sample size, time window to collect data, and the inclusion of computer interaction features.
	\item A discussion on the impact of applying this research to improve
the interaction of knowledge workers with their digital environment leading to an improvement in their productivity and well-being, besides a reflection on aspects that can be further improved in future studies.
    \end{itemize}

>>>>

%!TEX root = bioPrediction_main.tex
\section{Introduction}
\textit{`The most valuable asset of a 21st-century institution (whether 
business or non-business) will be its knowledge workers and their 
productivity'}~\cite{drucker1999knowledge}. Knowledge workers constantly face 
challenges, such as a high work fragmentation, continuous disruptions and 
distractions, highly complex and demanding tasks, and long working 
hours~\cite{gonzalez2004constant,mark2008cost,czerwinski04diary}. 
These challenges, amongst others, can lead to stress in the workplace.
Stress is an ever-growing concern as it can lead to
fatigue, burnout and various other illnesses, ultimately resulting in work 
absences and marked productivity 
losses~\cite{hockey1997stress,setz2010stress,wrs2010}.

Given the importance of understanding stress and its relationship and
effect on work, a number of studies have been conducted using a
variety of methods. Some have studied stress using invasive
techniques, such as measuring cortisol as a stress biomarker~\cite{piazza10} which are best suited
for laboratory environments.  Further research has shown the benefit of using
autonomic nervous system (ANS) activity by analyzing biometric signals
of the human body, such as blood pressure~\cite{kataoka00}.
These less invasive approaches can be used in longer-term field
studies, opening up the question of what stress
looks like in the wild.

In this paper, we build on this ability to non-invasively monitor stress
to investigate two questions: RQ1) how do knowledge
workers at work experience stress over time, and RQ2) can we predict
whether a knowledge worker is experiencing stress in the moment based
on biometric data. A better understanding of how stress is experienced (RQ1)
can help inform the design of workplaces to manage
and alleviate stress. An ability to predict stress in the moment (RQ2)
can enable the development of digital tools to avoid stress.

Our work builds upon previous work and extends it by performing an
eight-week study in the workplace with 14 knowledge 
workers. To the best of our knowledge, this is the 
longest in-situ study performed in a real-world 
environment of knowledge workers to analyze an extensive collection of biometric signals with experience samples examining real-time prediction of stress in the workplace~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015}.
Previous studies are predominantly controlled lab studies which used only a subset of the relevant biometric signals we captured, have been much more limited in duration, or are not focused on the workplace.

We collected biometric data as well as
experience samples on stress, and two related human aspects: focus and
awakeness. We collected participant reports on focus because the
challenges that contribute to stress levels can also make it hard for
knowledge workers to stay focused and the level of focus
can affect
productivity~\cite{mark2014bored}.  We collected participant reports
on awakeness because a lack of awakeness (sleepiness) can result in
undesired consequences on work~\cite{connor02}.  The participants in
our study, who perform various job functions for a research and
development group within a single large corporation, wore a novel, state-of-the-art
biometric armband sensor\footnote{Biovotion Everion
  sensor~\cite{everion}} with low invasiveness that captured heart-,
respiration- and skin-related measures. This modality was chosen to
ease longitudinal deployment in the field. Using machine learning, we
created classifiers and analyzed their ability to predict stress, focus, and awakeness levels. Considering all of these three human aspects helps us to consider whether one aspect might be easier to detect than another. If one aspect is easier to detect, it might serve
as a proxy or an indicator of the presence (or absence) of other aspects.

In our analysis, we identify several trends in day-to-day stress levels
that emerged over the course of our eight-week study.  The results of
our analysis also show that biometric signals collected from a single
minimally invasive sensor can be used to predict stress and its
related aspects accurately, with the abstract concept of focus
(predictably) being the hardest to detect. Our results further show
that knowledge workers' self-reported levels of stress, focus and
awakeness and their physiological manifestation and prediction can
vary substantially between individuals.
%Our results further show that overall combining biometric features with computer interaction information 
%results in an improvement
%over each of the training sets individually.
The main contributions of our work are:
\begin{itemize}
	\item A qualitative examination of how knowledge worker stress presents and fluctuates in the wild.
	\item The creation and analysis of measures for the automatic monitoring of knowledge workers' stress, focus and awakeneess in the workplace based on an eight-week field study with 14 office workers.
	%\item A thorough analysis of ways to increase performance, including the training sample size, time window to collect data, and the inclusion of computer interaction features.
	\item A discussion on the impact of applying this research to improve
the interaction of knowledge workers with their digital environment leading to an improvement in their productivity and well-being, besides a reflection on aspects that can be further improved in future studies.
    \end{itemize}

### 2020-02-04 UbiCompSubmission/bp_relatedWork.tex
addressing rebuttal bullets

Eye tracking technology~\cite{Bednarik2006,Crosby1990,Rodeghero2014} and brain activity~\cite{Ikutani2014,Siegmund2014} 
have been used in previous studies to analyze different tasks in an office environment.
Eye tracking has been used to analyze memory load and processing load by inspecting task-evoked pupillary response and pupil size~\cite{beatty82}. Similar studies have shown high correlation between pupil size and mental workload of subtasks~\cite{beatty82} and cognitive load~\cite{Klingner10}.
Brain activity has been associated with different mental states~\cite{Berger29} by analyzing specific frequency bands (alpha, beta, gamma, delta, and theta) using electroencephalography (EEG). The increase or decrease of some of these frequencies is correlated with
attentional demand and working memory load~\cite{Smith05,sterman93}.
In contrast to studies that use eye tracking or EEG, we focused on a less invasive technology that can be applied in a real world scenario.

>>>>

Eye tracking technology~\cite{Bednarik2006,Crosby1990,Rodeghero2014} and brain activity~\cite{Ikutani2014,Siegmund2014} 
have been used in previous studies to analyze different tasks in an office environment.
Eye tracking has been used to analyze memory load and processing load by inspecting task-evoked pupillary response and pupil size~\cite{beatty82}. Similar studies have shown high correlation between pupil size and mental workload of subtasks~\cite{beatty82} and cognitive load~\cite{Klingner10}.
Brain activity has been associated with different mental states~\cite{Berger29} by analyzing specific frequency bands (alpha, beta, gamma, delta, and theta) using electroencephalography (EEG). The increase or decrease of some of these frequencies is correlated with
attentional demand and working memory load~\cite{Smith05,sterman93}.
In contrast to studies that use eye tracking or EEG, we focused on a less invasive technology that can be applied in a real world scenario.
Similarly, we considered adding additional metrics such as Depression, Anxiety and Stress Scale (DASS) and Perceived Stress Scale (PSS)~\cite{Ferdous15} but opted for the ones presented in the paper after several discussions with experts in the area, and after piloting surveys to maxime participant compliance.

### 2020-02-04 UbiCompSubmission/bp_results_1.tex
addressing rebuttal bullets

To examine the performance of a general model for our participants, we trained three general models, one for focus, one for awakeness and one for stress. We roughly followed the same procedure as for the individual models. Due to the larger amount of data available in the general case, we used the more common random undersampling, which randomly selects elements in the majority class to exclude from the dataset, instead of random oversampling to balance the distribution of the dataset. The models were trained on the datasets of 13 of the 14 participants, and then evaluated on the dataset of the last, repeating this process for all 14 participants.

>>>>

To examine the performance of a general model for our participants, we trained three general models, one for focus, one for awakeness and one for stress. We roughly followed the same procedure as for the individual models. Due to the larger amount of data available in the general case, we used the more common random undersampling, which randomly selects elements in the majority class to exclude from the dataset, instead of random oversampling to balance the distribution of the dataset. The models were trained on the datasets of 13 of the 14 participants, and then evaluated on the dataset of the last (leave-one-group-out cross-validation), repeating this process for all 14 participants.

### 2020-02-04 UbiCompSubmission/bp_studyMethod.tex
addressing rebuttal bullets

\subsection{Participants}
We recruited 14 professionals via personal contacts from a large power and automation company. All participants work primarily in an office environment, though half of the participants spend at least 10\% (and up to 50\%) of their time in a laboratory environment. Office workers are a population that generalizes to a variety of contexts, and including part-time laboratory workers guarantees that our participants have varying work patterns that include different levels of computer usage, as well as different levels of activity in both individual and collaborative tasks.

>>>>

\subsection{Participants}
We recruited group of 14 professionals via personal contacts from a large power and automation company. The group is very diverse in terms of age, work experience, gender, and work responsibilities. All participants work primarily in an office environment, though half of the participants spend at least 10\% (and up to 50\%) of their time in a laboratory environment. Office workers are a population that generalizes to a variety of contexts, and including part-time laboratory workers guarantees that our participants have varying work patterns that include different levels of computer usage, as well as different levels of activity in both individual and collaborative tasks.
